Value-Suppressing Uncertainty Maps

### 2018-01-04 paper/figures.tex
Replace all VSUM with VSUP
Fleshed out discussion section.
More JH edits, prep for camera ready.

%Figure list:
% Vsum vs. traditional 2D heatmap vs. juxtaposed maps
% Chart of color bins w/r/t CIELAB threshold, with iconic maps at intervals
% Process figure
% Real examples with different color maps

@@

\newcommand{\conditionFig}{
	\begin{figure*}[t]
		\centering
		\includegraphics[width=0.95\textwidth]{allConditions.pdf}
		\caption{The 8 conditions from the identification experiment. Juxtaposed maps require a participants to make an error-prone connection between areas in two separate maps in order to make a decision that integrates value and uncertainty. Traditional bivariate maps integrate both value and uncertainty. VSUPs attempt to improve on traditional bivariate maps by reducing color resolution as uncertainty increases, discouraging conclusions based on noisy or imprecise data.}
		\label{fig:conditions}
	\end{figure*}
}

\newcommand{\taskOneFig}{
	\begin{figure}
		\centering
		\includegraphics[width=0.95\columnwidth]{task1.pdf}
		\caption{Accuracy results for the identification experiment. For examples of each condition, see \protect\figref{fig:conditions}. Juxtaposing two univariate maps for both value and uncertainty requires an error-prone search task for identification tasks. Continuous rather than discrete bivariate maps requires an error-prone color encoding and estimation task. Discrete bivariate maps, both VSUPs and otherwise, avoid these issues. The confidence intervals are bootstrapped 95\% CIs of trimmed means.}
		\label{fig:taskOne}
	\end{figure}
}

@@

	\centering
	\includegraphics[width=\columnwidth]{airline-vsum}
	\vspace{-15px}
	\caption{Flight delay data encoded with a VSUP.}
	\label{fig:airlineVsum}
\end{figure}
}

@@

	\centering
	\includegraphics[width=\columnwidth]{viral-vsum.pdf}
	\vspace{-15px}
	\caption{Variability for 12 different viral populations of SIV, encoded with a VSUP. By suppressing uncertain information, the VSUP version removes the apparent spike in value at the second column, which is an artifact of sampling error due to poor coverage in that region. It also suppresses apparent inter-species variability in the solid orange spike at location ~9,500, which, again, is a function of poor coverage rather than a genuine per-strain difference.  }
	\label{fig:viralVsum}
\end{figure}
}

@@

		\centering
		\includegraphics[width=\columnwidth]{polling-vsum.pdf}
		\vspace{-15px}
		\caption{Polling data prior to the 2016 U.S. Presidential Election, encoded with a VSUP. }
		\label{fig:pollingVsum}
	\end{figure}
}

>>>>

%Figure list:
% Vsup vs. traditional 2D heatmap vs. juxtaposed maps
% Chart of color bins w/r/t CIELAB threshold, with iconic maps at intervals
% Process figure
% Real examples with different color maps

@@

\newcommand{\conditionFig}{
	\begin{figure*}[t]
		\centering
		\includegraphics[width=0.95\textwidth]{allConditions.pdf}
		\caption{The 8 conditions from the identification experiment. Juxtaposed maps require participants to make an error-prone connection between areas in two separate maps in order to make a decision that integrates value and uncertainty. Traditional bivariate maps integrate both value and uncertainty. VSUPs attempt to improve on traditional bivariate maps by reducing color resolution as uncertainty increases, discouraging conclusions based on noisy or imprecise data.}
		\label{fig:conditions}
	\end{figure*}
}

\newcommand{\taskOneFig}{
	\begin{figure}[h!]
		\centering
		\includegraphics[width=0.95\columnwidth]{task1.pdf}
		\caption{Accuracy results for the identification experiment. For examples of each condition, see \protect\figref{fig:conditions}. Juxtaposing two univariate maps for both value and uncertainty requires an error-prone search task for identification tasks. Continuous rather than discrete bivariate maps requires an error-prone color encoding and estimation task. Discrete bivariate maps, both VSUPs and otherwise, avoid these issues. The confidence intervals are bootstrapped 95\% CIs of trimmed means.}
		\label{fig:taskOne}
	\end{figure}
}

@@

	\centering
	\includegraphics[width=\columnwidth]{airline-vsup}
	\vspace{-15px}
	\caption{Flight delay data encoded with a VSUP.}
	\label{fig:airlineVsup}
\end{figure}
}

@@

	\centering
	\includegraphics[width=\columnwidth]{viral-vsup.pdf}
	\vspace{-15px}
	\caption{Variability for 12 different viral populations of SIV, encoded with a VSUP. By suppressing uncertain information, the VSUP version removes the apparent spike in value at the second column, which is an artifact of sampling error due to poor coverage in that region. It also suppresses apparent inter-species variability in the solid orange spike at location ~9,500, which, again, is a function of poor coverage rather than a genuine per-strain difference.  }
	\label{fig:viralVsup}
\end{figure}
}

@@

		\centering
		\includegraphics[width=\columnwidth]{polling-vsup.pdf}
		\vspace{-15px}
		\caption{Polling data prior to the 2016 U.S. Presidential Election, encoded with a VSUP. }
		\label{fig:pollingVsup}
	\end{figure}
}

### 2018-01-04 paper/main.tex
More JH edits, prep for camera ready.

Given that bivariate maps, for reasons of either practicality or perceptual fidelity, have only a limited budget of outputs, we adapt an insight from Dunn~\cite{dunn1989dynamic} that these limited categories ought to be assigned with regards to their importance to analysis tasks, rather than uniformly. Our Value-Suppressing Uncertainty Palettes apply this insight to data with uncertainty. Frequently, uncertain information ought to be weighted less heavily in a decision-making process, or at least have less prominence in an analysis, than information with high certainty.

@@

JavaScript code for generating VSUPs for use in D3~\cite{bostock2011d3} charts is available at \url{https://github.com/uwdata/spatial-sampling}.

@@

Using a public data set from Wikipedia~\cite{wiki} of the statewide polls that were conducted closest to the 2016 U.S. presidential election, we created a choropleth map showing which candidate was leading in the polls. We calculated uncertainty as the size of the lead in terms of margin of error (lead/margin). The AP style guide for reporters suggests that a candidate leading by more than two margins of error is said to be ``significantly ahead'' in the polls~\cite{french1988associated}. As the lead becomes closer to the margin of error, there is more doubt about the electoral outcome.

@@

Experimental materials, including data tables and stimuli generation code, are available at \url{https://github.com/uwdata/spatial-sampling}.

@@

For the prediction task, we gave participants the rules of a game similar to Battleship. Greis \ea~\cite{greis2016decision} employ these game-like experimental tasks to assess how different visual designs communicate uncertainty information, which can be abstract or complex, to the general audience. In our task, the participant and a (fictional) adversary have to place tokens representing ships on a $5$x$5$ spatial grid, with the expectation that certain squares will be hit by missiles. Players have to place all their tokens before continuing. The objective is to minimize the number of your own ships that are hit. In our task, participants were given a map representing the predictions of missile strikes in each location on the grid. The \emph{value} component was the ship's \emph{danger} if placed on the square. The \emph{uncertainty} component was the confidence in this prediction. Other studies of uncertainty representation, such as in Cox et al.~\cite{cox2013visualizing}, have used "prediction + prediction uncertainty" stimuli to elicit differences in decision-making between visualizations of uncertainty.

@@

We selected this task in order to promote risk-averse behavior. Tversky \& Kahneman~\cite{tversky1985framing} illustrate that framings in terms of gains or losses produce reliably different outcomes. In particular, there is greater perceived value in avoiding large losses as opposed to striving for a large gain~\cite{kahneman1979prospect}. Our results from the prior study indicate that discrete, non-juxtaposed maps outperformed the other bivariate maps we selected, so we limited our study to only 4 types: square and wedge bivariate maps. While other map types might produce different patterns of decision-making (for instance, the juxtaposed map might encourage participants to ignore uncertainty information altogether), their low accuracy for our previous information fusion tasks led us to discard them, as it would be difficult to distinguish between different patterns of predictions caused by the design, and different patterns caused by simply misreading the heatmap. Having fewer conditions also afforded a within-subjects design that controlled for the variation in interpersonal differences in strategies and risk-aversion, while limiting the potential effects of learning and fatigue on predictions from large numbers of stimuli.

@@

VSUPs are in a middle ground between sparse but simple, and dense but complex presentations of uncertainty. At high levels of uncertainty, VSUPs have only one output color, and so explicitly prevent any decoding of value. This narrowing of options is reflected not only in the quantization scheme, but also in the shape of the legend, narrowing as the uncertainty increases. We selected this design to promote \emph{risk-averse} decision-making. That is, people should be hesitant to make strong predictions or decisions about values when uncertainty is high. Our prediction experiment, drawing from Kahneman \& Tversky's prospect theory~\cite{kahneman1979prospect}, was further designed to elicit avoidance of uncertain data, by framing the predictions in terms of minimizing losses, rather than maximizing gains. This framing was reflected in participants mostly making safe and certain guesses (low danger, high certainty). Where the differences in decision-making occurred were where the participants had exhausted the set of safe and certain locations, and were forced to value either certainty or expected value. A risk seeker would choose uncertain locations if the reward were high enough, whereas a risk-avoider would select locations with less uncertainty, even with a lower payoff. Participants using VSUPs were likelier to make these risk-avoiding decisions than when they were using traditional bivariate maps. Beyond VSUPs, this result indicates that the designers of visualizations can have fine-grained control over the types of decision-making their visualizations promote. Visualization designers should consider the not just fidelity to the underlying dataset, but also the utility of the decisions their designs support.
%\sizeFig

@@

VSUPs assume that uncertain values ought to have less visual impact that highly certain values. VSUPs therefore act both as a kind of filtering device as well as a bivariate representation. This assumption holds in scenarios like the polling dataset (\figref{fig:pollingVsup}), where the expectation is that highly uncertain data is unreliable or should otherwise be downweighted in analysis. If the analyst has a different interpretation of uncertainty, or wishes to quickly and orthogonally analyze the distributions of uncertainty and value, other strategies, such as juxtaposed maps, may be more appropriate. VSUPs are designed for the \emph{integration} of value and uncertainty. Designers should take care in considering when and how this integration is desirable.

>>>>

Given that bivariate maps, for reasons of either practicality or perceptual fidelity, have only a limited budget of outputs, we adapt an insight from Dunn~\cite{dunn1989dynamic} that these limited categories ought to be assigned with regards to their importance to analysis tasks, rather than uniformly. Our Value-Suppressing Uncertainty Palettes apply this insight to data with uncertainty. Frequently, uncertain information ought to be weighted less heavily in a decision-making process, or have less prominence in an analysis, than information with high certainty.

@@

JavaScript code for generating VSUPs for use in D3~\cite{bostock2011d3} charts is available at \url{https://github.com/uwdata/vsup}.

@@

Using a public data set from Wikipedia~\cite{wiki} of the statewide polls that were conducted closest to the 2016 U.S. presidential election, we created a choropleth map showing which candidate was leading in the polls. We calculated uncertainty as the size of the lead in terms of margin of error (lead/margin). The AP style guide for reporters suggests that a candidate leading by more than two margins of error is said to be ``significantly ahead'' in the polls~\cite{french1988associated}. As the lead becomes closer to the margin of error, there is more doubt about the electoral outcome, and smaller leads are less definitive.

@@

Experimental materials, including data tables and stimuli generation code, are available at \url{https://github.com/uwdata/vsup-paper}.

@@

For the prediction task, we gave participants the rules of a game similar to Battleship. Greis \ea~\cite{greis2016decision} employ these game-like experimental tasks to assess how different visual designs communicate uncertainty information, which can be abstract or complex, to the general audience. In our task, the participant and a (fictional) adversary have to place tokens representing ships on a $5$x$5$ spatial grid, with the expectation that certain squares will be hit by missiles. Players have to place all their tokens before continuing. The objective is to minimize the number of your own ships that are hit. In our task, participants were given a map representing the predictions of missile strikes in each location on the grid. The \emph{value} component was the ship's \emph{danger} if placed on the square. The \emph{uncertainty} component was the confidence in this prediction. Other studies of uncertainty representation, such as in Cox et al.~\cite{cox2013visualizing}, have used ``prediction + prediction uncertainty'' stimuli to elicit differences in decision-making between visualizations of uncertainty.

@@

We selected this task in order to promote risk-averse behavior. Tversky \& Kahneman~\cite{tversky1985framing} illustrate that framings in terms of gains or losses produce reliably different outcomes. In particular, there is greater perceived value in avoiding large losses as opposed to striving for a large gain~\cite{kahneman1979prospect}. Our results from the prior study indicate that discrete, non-juxtaposed maps outperformed the other bivariate maps we selected, so we limited our study to only 4 types: square and wedge bivariate maps. While other map types might produce different patterns of decision-making (for instance, the juxtaposed map might encourage participants to ignore uncertainty information altogether), their low accuracy for our previous information fusion tasks led us to discard them, as it would be difficult to distinguish between different patterns of predictions caused by the design, and different patterns caused by simply misreading the heatmap. Having fewer conditions also afforded a within-subjects design that controlled for the variation in interpersonal differences in strategies and risk-aversion, while limiting the potential effects of learning and fatigue from large numbers of stimuli.

@@

VSUPs are in a middle ground between sparse but simple, and dense but complex presentations of uncertainty. At high levels of uncertainty, VSUPs have only one output color, and so explicitly prevent any decoding of value. This narrowing of options is reflected not only in the quantization scheme, but also in the shape of the legend, narrowing as the uncertainty increases. We selected this design to promote \emph{risk-averse} decision-making. That is, people should be hesitant to make strong predictions or decisions about values when uncertainty is high. Our prediction experiment, drawing from Kahneman \& Tversky's prospect theory~\cite{kahneman1979prospect}, was further designed to elicit avoidance of uncertain data, by framing the predictions in terms of minimizing losses, rather than maximizing gains. This framing was reflected in participants mostly making safe and certain guesses (low danger, high certainty). Where the differences in decision-making occurred were where the participants had exhausted the set of safe and certain locations, and were forced to value either certainty or expected value. A risk seeker would choose uncertain locations if the reward were high enough, whereas a risk-avoider would select locations with less uncertainty, even with a lower payoff. Participants using VSUPs were likelier to make these risk-avoiding decisions than when they were using traditional bivariate maps. Beyond VSUPs, this result indicates that the designers of visualizations can have fine-grained control over the types of decision-making their visualizations promote. Visualization designers should consider not just fidelity to the underlying dataset, but also the decision strategies their designs support.
%\sizeFig

@@

VSUPs assume that uncertain values ought to have less visual impact than highly certain values. VSUPs therefore act both as a kind of filtering device as well as a bivariate representation. This assumption holds in scenarios like the polling dataset (\figref{fig:pollingVsup}), where the expectation is that highly uncertain data is unreliable or should otherwise be downweighted in analysis. If the analyst has a different interpretation of uncertainty, or wishes to quickly and orthogonally analyze the distributions of uncertainty and value, other strategies, such as juxtaposed maps, may be more appropriate. VSUPs are designed for the \emph{integration} of value and uncertainty. Designers should take care in considering when and how this integration is desirable.
